---
title: "Effect Sizes for single-subject design studies using Bayesian hierarchical models"
description: |
  In this post, a method for leveraging Bayesian methodology to estimate individual effect sizes in single-subject designs is discussed in detail. 
author:
  - name: Rob Cavanaugh
date: 12-28-2020
categories:
  - stats
  - aphasia treatment
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_depth: 2
draft: true
bibliography: citations.bib

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
In _A Systematic Appraisal of Effect Sizes in Aphasia Single-Case Design via Simulation_ [@cavanaugh_2020], we compared different effect size measures that have been used in the aphasia single-case design literature using simulation. We included an effect size method which we used in one of [our](https://lrcl.pitt.edu){target="_blank" rel="noreferrer} recent aphasia treatment studies [@evans2020] and are using in another (currently in preparation). In essence, this approach models treatment probe performance over the course of treatment using an interrupted time series approach [@huitema2000] and then samples the posterior distribution to estimate change in the number of items treated. This post will walk through this approach in detail, using a sample of simulated data from our simulation study.

## Setup
```{r packages, warning=F, message=F, echo = T}
# load packages
library(here)
library(tidyverse)
library(brms)
library(tidybayes)
library(bayesplot)
```


```{r read, warning=F, message=F, echo = T}
# read in data
df <- read_csv(here('data', 'shiny_data.csv'))  %>%
  mutate(sub_id = as.factor(sub_id),
         phase = as.factor(phase))

df[10,]
```

